### 2015-04-08 PLDI15_Response/response_notes.txt

==============================================================================
Common Concerns
==============================================================================

Most important : Complexity of Contracts
—————————————————————————————————————————

(Reviewer 5; main reason for rejection) - It isn't clear that the proposed
solution is any less complicated than reasoning directly about consistency; the
evaluation does not actually evaluate this hypothesis, which is the main
motivation for the paper.

(Reviewer 5) The contract language is complex and provides an even richer set
of choices to the user than just choosing an appropriate consistency level---it
isn't clear to me that this is an improvement for the user (e.g., the example
for withdraw and getBalance on page 4 seems to require a deep understanding of
the requirements). If a realistic scenario is just three totally-ordered
consistency levels (as per this paper's evaluation), then it seems simpler to
choose the level directly rather than try to specify a contract. This problem
might be mitigated if it is expected than an application might be run on
multiple different data stores with very different sets of consistency levels;
is this the case here?

(Reviewer 2) "As the author's acknowledge, the contracts language available to
the programmer makes for verbose and low-level annotation requirements.
Moreover, there is scant evidence that writing these annotations is any easier
to get right than picking a consistency level in the first place."

(Reviewer 3) “One of my concern of the paper is the expression of contracts is
still complex. A example is the getBalance contract Page 4. While the current
contract language is expressive, it is also a little complex if the purpose is
just to specify the consistency level of a certain data structure."

(Reviewer 4) “One of the problems with new languages or new language extensions
is that they usually do not get adopted by programmers. A programmer would take
the time to learn a new language only if he feels the time spent on learning
will pay off eventually. I am not sure programmers will be willing to adopt
this extension."

** Start Response **

— Programmers write contracts for the given abstract system model, which admits
all the inconsistencies associated with eventual consistency. However, the
programmer only needs to reason about inconsistencies in this abstract model
not the intricacies of the actual store model, which are often left
unspecified. The advantage of Quelea contract is that programmer-written
contracts are automatically mapped onto the store consistency level.

** End Response **

Delta
—————

(Reviewer 3) Another issue of this paper is the delta between this paper and
the POPL paper[ Burckhardt et al. 2014] on RDT. In the POPL paper, the contract
was used as languages for verification tool while in this paper it is used to
generate distributed protocol code. There are certainly differences between
these two usage model, but the prior work does make this paper a little
incremental.

(Reviewer 4) The increment of this work over [Bailis 2013b] is not very large.
The previous work (and I don’t know if it originates from the same authors)
already proposed separation of concerns and implemented an initial scheme. The
current paper seems to extend it with a language and the use of the theorem
prover. As a non-expert, this seems to me like a nice step but not a very large
one.

Execution Time
——————————————

(Reviewer 4) Can you say how much time the execution of the theorem prover
takes? If the consistency constraints become more complex will the theorem
prover still be able to determine the appropriate levels of support
efficiently?

(Reviewer 1) I believe the compile time can still be a show stopper. Given
these applications can change a lot, isn't the long compilation time due to the
SMT solver an issue?

===============================================================================
Reviewer specific concerns
===============================================================================

Reviewer 1
——————————

I think also comparing to a dynamic strategy for enforcing contacts (mentioned
on page 5 as an alternative), would make the case stronger. Such strategy can
take advantage from transaction buffer in each replica. Can you comment?

-I believe the compile time can still be a show stopper. Given these
applications can change a lot, isn't the long compilation time due to the SMT
solver an issue?

-In your experiments did you consider the shim layer overhead? Did you use the
same number of nodes in all tests? I mean was it the case that "shim layer
nodes" + "other nodes" in Q runs was always equal to "all the nodes" in non-Q
runs? Also, how does the shim layer grow (number of nodes) as the whole system
grows?

-Are all the benchmarks written from scratch or did you convert some existing
one to QUELEA? My impression is the former is correct.

Reviewer 2
——————————

The paper doesn't provide much discussion of the limitations of the
approach---either positive or negative. For example, are there notions of
consistency that fundamentally would not fit into this model?  The end of the
paper mentions that it doesn't handle notions that depend on wall-clock time as
distinct from session order. It would be interesting to know whether notions
such as bounded staleness could be readily accomodated or not.

As the author's acknowledge, the contracts language available to the programmer
makes for verbose and low-level annotation requirements. Moreover, there is
scant evidence that writing these annotations is any easier to get right than
picking a consistency level in the first place.

The system model requires the clients to perform all the actual computation on
the data-store, including merging and summarizing. In a production setting this
would likely prove infeasible, as it relies on strong assumptions of the
clients' relationships, including identical running code and complete trust.
The possibility that a crashed operation would be durably added to the log,
only to break when a different client attempts to reduce over the log much
later, is not friendly.

** Start Response **

The RDT specification language that we have described does to some extent gaurd
against malicious/incorrect code. In particular, the operations on RDT are
described as reductions over the history, and in particular is side-efffect
free (which is ensured by the type of the operation -- section 3.1). While it
is possible that the reduction can end up not terminating, it cannot do any
more harm.

Crashed operation durably added to log?? By definition, an operation is added
to log only if it is successful; if the operation procudes a result, then it
has succeeded. Hence a crashed operation will not be added to log.

Summarization operation checks whether the summarized history is smaller than
the input history. Otherwise, the summarization operation is considered
incorrect and rejected.

** End Response **

Minor comment: Footnote 3 discusses how you "encode" transitive closure in
first-order logic. I didn't understand how this could ever be useful. Given the
constraints

	R(a,b) => R'(a,b)

	R'(a,b) /\ R'(b,c) => R'(a,c)

it seems that R' could be the total relation. How would this ever be useful?

** Start Response **

R+ could be a total relation, but the R+ is still useful because it is not
required to be a total relation. For example, consider causal
consistency (CC) contract which is defined as:

 hbo \subseteq vis

where hbo = (soo U vis)+. Because of our lenient encoding of “+", hbo can be
any superset of  transitive closure of (soo U vis). Nevertheless, all we can
assume about hbo is that it contains the  transitive closure of (soo U vis);
there is no proof of membership for other elements in hbo.  Consequently, by
asserting hbo \subseteq vis, the strongest guarantee that causal consistency 
contract is offering is that the transitive closure of (soo U vis) is contained
in vis. An application  contract (\psi) which requires more than the transitive
closure of (soo U vis) to be contained in vis  cannot be satisfied under CC
because the following implication is simply not valid even under our lenient
definition of “+“ :

 \Delta |- (hbo \subseteq vis) => \psi

\Delta, as usual, denotes the conjunction of basic axioms about primitive
relations.

** End Response **

Reviewer 3
——————————

I am wondering if there are more intuitive and succinct ways to provide a
higher level abstraction to define these contracts? One example is the "atomic"
directives in Java. Can we have something similar to that to allow easier
declaration of consistency levels?

Reviewer 5
——————————

The evaluation shows that using the weakest levels possible while still
preserving correctness does give a performance improvement, but this isn't a
surprising result. I would be more interested in seeing whether using QUELEA
leads to fewer mis-classifications than choosing levels directly, which speaks
to the main motivation of this work.

This paper finds the correct consistency level by succesively testing the
levels in order of strength in order to find the weakest one that satisfies the
contract. Is it expected that most stores would have only a few levels in total
order? If there are many levels in a partial order, is finding the correct
level still tractable? If the levels are partially ordered, how is the "best"
level defined?

** Start Response **

Most stores that are available only tend to expose a few levels, which can
indeed be partially ordered. An example is classification from [[Doug Terry’s
Session guarantees for weakly cons]stent replicated data]] which introduced 4
(incomparable) session guarantees, and the consistency level for an operation
is any combination of the 4 session guarantees. The classification here, at
worst, requires 4! = 24 comparisons (SMT validity tests). However, each
comparison is very cheap; on average, in our evaluation, each comparison took
11.5 ms. Hence, the solution remains scalable.

“Best Level?” Since the classification is parameterised, when different
incomparable consistency levels fit the bill, the RDT implementor can either
choose one of the levels, or perhaps choose a stronger level that encapsulates
both. This decision is left to the RDT implementor.

** End Response **

### 2015-04-08 store-opsem.tex
Fixed typos.
Applied changes from paper to techrep

The state of an operational execution ($\E$) is a tuple $\Exec$ where
$\EffSoup$ is a set of effects, and $\visZ$, $\soZ$, $\sameobjZ$
$\subseteq$ $\EffSoup \times \EffSoup$ are \emph{visibility},
\emph{session order}, and \emph{same object} relations over effects,
respectively. We define an effect ($\eff$) as a tuple
$(s,i,op,v)$, which records the fact that $i^{th}$ action in session
with {\sf SessID} $s$, which is an operation $op$ on the replicated
object, has been successfully executed on some replica yielding a
return value $v$. Note that the combination of $s$ and $i$ uniquely
identifies the effect. Session order relation ($\soZ$) relates effects
generated by the same session.  An effect $\eff = (s,i,op,v)$ is said
to precede another effect $\eff' = (s',i',op',v')$ in session order if
and only if $s'=s$ and $i'\ge i$. Since we only consider one
replicated object in our formalization, the $\sameobjZ$ relation
relates every pair of effects in the effect soup ($\EffSoup$). An
effect generated at a replica becomes visible at rest of the replicas
eventually.  If we denote the effect generated by the operation $op$
as $\eff_{op}$, then $\Theta(r) \times \{\eff_{op}\} ~\subseteq~
\visZ$. Often, in our formalization, we use $\visZ$ and $\soZ$ binary
relations to obtain a set of effects visible to a given effect $\eff$,
or set of effects that precede a given effect $\eff$ in the session
order. As a syntactic convenience, whenever $R$ is a binary relation,
we write $R(\eff)$ to denote the set of all $\eff'$ such that
$(\eff,\eff') \in R$.  Conversely, we write $R^{-1}(\eff)$ to denote
the set of all $\eff'$ such that $(\eff',\eff) \in R$.

@@

\begin{definition}
  A store $\Theta$ is said to be causally consistent under an execution
  $\E=\Exec$ if and only if:
  \begin{smathpar}
  \begin{array}{cr}
    \hspace*{-0.5in}\forall (r\in{\sf dom}(\Theta)). \forall (\eff \in
      \Theta(r)). & \\
    \hspace*{0.3in}\forall (a\in \EffSoup). \hbo{a}{\eff} \Rightarrow a 
      \in \Theta(r) & H\npp \\
  \end{array}
  \end{smathpar}
  Where, $\hboZ = (\Rvis \cup \Rso)^{+}$
\end{definition}

@@

\begin{corollary}[Soundness]
  \label{thm:soundness}
  For every well-formed execution state $\E$, for every store $\Theta$
  that is causally consistent under $\E$, for every contract class $\tau
  \in \{{\sf ec},{\sf cc},{\sf sc}\}$, and for every consistency
  contract $\cv$ in the contract class $\tau$, If:
  \begin{smathpar}
  (\E, \Theta, \langle s,i,op_\tau :: 
  \sigma \rangle \pll \Sigma) \xrightarrow{\eff} (\E', \Theta',
  \langle s,i+1,\sigma \rangle \pll \Sigma)
  \end{smathpar}
  then (\romannumeral 1) $\E'$ is well-formed, (\romannumeral 2)
  $\Theta'$ is causally consistent under $\E'$, and (\romannumeral 3)
  $\E' \models \cv[\eff/\cureff]$
\end{corollary}
\begin{proof}
  Follows from Theorems \ref{thm:classification-sound},
  \ref{thm:local-state-is-cc}, and \ref{thm:weak-soundness}.
  \qed
\end{proof}

>>>>

The state of an operational execution ($\E$) is a tuple \Exec, where
$\EffSoup$ is a set of effects, and $\visZ$, $\soZ$, $\sameobjZ$
$\subseteq$ $\EffSoup \times \EffSoup$ are \emph{visibility},
\emph{session order}, and \emph{same object} relations over effects,
respectively. We define an effect ($\eff$) as a tuple
$(s,i,op,v)$, which records the fact that $i^{th}$ action in session
with {\sf SessID} $s$, which is an operation $op$ on the replicated
object, has been successfully executed on some replica yielding a
return value $v$. Note that the combination of $s$ and $i$ uniquely
identifies the effect. Session order relation ($\soZ$) relates effects
generated by the same session.  An effect $\eff = (s,i,op,v)$ is said
to precede another effect $\eff' = (s',i',op',v')$ in session order if
and only if $s'=s$ and $i'\ge i$. Since we only consider one
replicated object in our formalization, the $\sameobjZ$ relation
relates every pair of effects in the effect soup ($\EffSoup$). An
effect generated at a replica becomes visible at rest of the replicas
eventually.  If we denote the effect generated by the operation $op$
as $\eff_{op}$, then $\Theta(r) \times \{\eff_{op}\} ~\subseteq~
\visZ$. Often, in our formalization, we use $\visZ$ and $\soZ$ binary
relations to obtain a set of effects visible to a given effect $\eff$,
or set of effects that precede a given effect $\eff$ in the session
order. As a syntactic convenience, whenever $R$ is a binary relation,
we write $R(\eff)$ to denote the set of all $\eff'$ such that
$(\eff,\eff') \in R$.  Conversely, we write $R^{-1}(\eff)$ to denote
the set of all $\eff'$ such that $(\eff',\eff) \in R$.

@@

\begin{definition}
  A store $\Theta$ is said to be causally consistent under an execution
  $\E=\Exec$ if and only if:
  \begin{smathpar}
  \begin{array}{cr}
    \hspace*{-0.5in}\forall (r\in{\sf dom}(\Theta)). \forall (\eff \in
      \Theta(r)). & \\
    \hspace*{0.3in}\forall (a\in \EffSoup). \hbo{a}{\eff} \Rightarrow a 
      \in \Theta(r) & H\npp \\
  \end{array}
  \end{smathpar}
  Where, $\hboZ = (\Rvis \cup \sooZ)^{+}$
\end{definition}

@@

\begin{corollary}[Soundness]
  \label{thm:soundness}
  For every well-formed execution state $\E$, for every store $\Theta$
  that is causally consistent under $\E$, for every contract class $\tau
  \in \{{\sf ec},{\sf cc},{\sf sc}\}$, and for every consistency
  contract $\cv$ in the contract class $\tau$, If:
  \begin{smathpar}
  (\E, \Theta, \langle s,i,op_\tau :: 
  \sigma \rangle \pll \Sigma) \xrightarrow{\eff} (\E', \Theta',
  \langle s,i+1,\sigma \rangle \pll \Sigma)
  \end{smathpar}
  then (\romannumeral 1) $\E'$ is well-formed, (\romannumeral 2)
  $\Theta'$ is causally consistent under $\E'$, and (\romannumeral 3)
  $\E' \models \cv[\eff/\cureff]$
\end{corollary}
\begin{proof}
  Follows from Theorems \ref{thm:classification-sound},
  \ref{thm:local-state-is-cc}, and \ref{thm:weak-soundness}.
  \hfill \qed
\end{proof}
